Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-12 Manuscript_Nater_etal.qmd
Post-merge clean-up

Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001]. Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]). While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space. However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space. Also, support for density dependence in willow ptarmigan has been mixed. @myrberget1988 observed no change in productivity despite a 50% decrease in abundance. Conversely, @pedersen2004 found strong overwinter negative density dependence and they posited that immigration was the vital rate that responded to changes in density. Low survival rates appeared to have a one-year lag following abundance peaks. This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 41 areas. Future work should explore different time-lags for density and population growth and the possibility for non-linear effects.

@@

<!--# They obviously use a lesser breed of bird dog up that way.Probably use more german shorthairs or vislas. :-)  -->

>>>>

Density dependence is an important process to understand in ecology but more so for exploited species like willow ptarmigan [@andrewartha1954; @sandercock_etal2011; @aanes2002; @willebrand2001]. Negative density dependence has been found in numerous gallinaceous birds such as northern bobwhites *Colinus virginianus* [@mcconnell2018], *Perdix perdix* [@bro2003], and wild turkeys *Meleagris gallopavo* [@mcghee2007]). While we found the highest population growth rates in areas with relatively low density, and relatively low growth rates in high-density areas [@fig-paramCorrs A], our analysis did not provide evidence for a strong relationship between density and growth rates across space. However, this is unsurprising as we did find substantial spatial differences in demography, which suggests that carrying capacity too varies over space. Also, support for density dependence in willow ptarmigan has been mixed. @myrberget1988 observed no change in productivity despite a 50% decrease in abundance. Conversely, @pedersen2004 found strong overwinter negative density dependence and they posited that immigration was the vital rate that responded to changes in density. Low survival rates appeared to have a one-year lag following abundance peaks. This pattern could be a result of negative density dependence that went undetected when looking at the ensemble of samples across the 41 areas. Negative density dependence has been shown in willow ptarmigan populations which may explain explain reduced survival following years with elevated densities [@pedersen2004].Future work should explore different time-lags for density and population growth and the possibility for non-linear effects.

<!--# Not happy with the DD para yet but I can work on Jan 11 -->

@@

<!--# They obviously use a lesser breed of bird dog up that way.Probably use more german shorthairs or vislas. :-)  -->
<!-- Funnily enough, I have been suggesting to Erlend that we should have a master student to analyze the effect of dog breed on detection probability :-P -->
